```{r knitsetup, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=FALSE}
opts_knit$set(base.dir='./', fig.path='', out.format='md')
opts_chunk$set(prompt=TRUE, comment='', results='markup')
# See yihui.name/knitr/options for more Knitr options.
##### Put other setup R code here

# end setup chunk
```
# Workshop 2 - Part 3
## Simple Stat Summaries

--- 

Now that we've read data in and exported data out of R, let's run some simple stat summaries.

## Some simple statistics

The `summary()` function is a quick simple way to get basic summary statistics on every variable in a dataset.

```{r}
data.csv <- read.csv(file="C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01_comma.csv")
summary(data.csv)

```

Let's make a histogram of the BMI's at PRE

```{r}
data.csv$bmiPRE <- (data.csv$WeightPRE*703)/((data.csv$Height*12)**2)
data.csv$bmiPOST <- (data.csv$WeightPOST*703)/((data.csv$Height*12)**2)
hist(data.csv$bmiPRE)
```

There is a typo, so let's fix the Height typo for subject 18. It is currently entered as 2.6 and should be 5.6. After fixing it we will update the BMI calculations and then replot the histogram.

We will also overlay a density curve.

```{r}
data.csv[18,"Height"] <- 5.6
data.csv$bmiPRE <- (data.csv$WeightPRE*703)/((data.csv$Height*12)**2)
data.csv$bmiPOST <- (data.csv$WeightPOST*703)/((data.csv$Height*12)**2)
hist(data.csv$bmiPRE, freq=FALSE)
lines(density(data.csv$bmiPRE))
```

Let's also make a quick scatterplot of BMI at PRE and POST and we'll overlay a linear best fit line using the `lm()` function and a non-parametric smoothed line using the `lowess()` function. We'll wrap the linear fit results with the `abline()` line function to overlay the best fit line and we'll use the `lines()` function to overlay the smoothed line.

```{r}
plot(data.csv$bmiPRE, data.csv$bmiPOST, "p")
abline(lm(data.csv$bmiPOST ~ data.csv$bmiPRE), col="red")
lines(lowess(data.csv$bmiPRE, data.csv$bmiPOST), col="blue")

```



