```{r knitsetup, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=FALSE}
opts_knit$set(base.dir='./', fig.path='', out.format='md')
opts_chunk$set(prompt=TRUE, comment='', results='markup')
# See yihui.name/knitr/options for more Knitr options.
##### Put other setup R code here

# end setup chunk
```
# Workshop 2 - Part 2
## Import, Export External Data

--- 

## Example Datafiles

I have created a simple dataset for us to work with. The files are available via Dropbox at [https://www.dropbox.com/sh/vlo5bzrl5ayo1bk/AADD0WieyuEdyGwiveuCoRr-a?dl=0](https://www.dropbox.com/sh/vlo5bzrl5ayo1bk/AADD0WieyuEdyGwiveuCoRr-a?dl=0)

```
The following files are in this folder:
02/18/2016  03:22 PM             2,513 Dataset_01.sav
09/13/2013  02:25 PM            30,208 Dataset_01.xls
09/07/2013  05:44 PM            10,338 Dataset_01.xlsx
09/13/2013  02:26 PM               567 Dataset_01_comma.csv
09/13/2013  02:26 PM               567 Dataset_01_tab.txt
09/07/2013  05:45 PM            30,022 NRSG736_01Handout.docx
```

These data files include:
* XLS - the older 1997-2003 EXCEL file format
* XLSX - the newer EXCEL file workbook format
* CSV - a comma delimited dataset
* xxx_tab.TXT - a TAB delimited dataset
* SAV - SPSS file format

We will work through importing each of these file types.

---

## The dataset description

The dataset we'll be working with here has 8 variables and 20 cases (or 20 subjects).

```{r echo=FALSE}
Dataset_01_comma <- read.csv(file="C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01_comma.csv")
kable(Dataset_01_comma, format="html", align='r')
```

```{r echo=FALSE, results="asis"}
# IGNORE FOR NOW - xtable options
#Try xtable - see below
# xt <- xtable(Dataset_01_comma)
# print(xt, type="html")
```

## IMPORT a basic CSV (comma delimited) or TAB (TAB delimited) text data file

Let's import a simple data file in CSV (comma delimited) format. This is the most common simple format used for exchanging data between different software/data handling packages.

In basic R the following functions are useful for reading in or importing in datasets:
* `read.table()`
* `read.csv()` 
* `read.delim()`

For each of these files you need to specify the data file, and possibly also supply whether the dataset has a header row, what the separator character is and more options. To see complete details see `help(read.table)`

For each example below, I am providing the specific directory path. However, if these datafiles are located in your current working directory you can just type the filename without the full path.

### Option 1 - use `read.table()`

The basic code is `read.table(file="xxx", header=TRUE, sep=",")` but these options can be changed as needed if your file does not have a header row or if your file has a separator other than a comma - some separators can be spaces or tabs.

```{r}
data.rt <- read.table(file="C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01_comma.csv",
                      header=TRUE,sep=",")
data.rt
```

### Option 2 - use `read.csv()`

Option 2 assumes that there is a header row and the datafile uses a comma separator.

```{r}
data.csv <- read.csv(file="C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01_comma.csv")
data.csv
```

### Option 3 - use `read.delim()`

Option 3 assumes that there is a header row and the datafile uses a TAB separator.

```{r}
data.tab <- read.delim(file="C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01_tab.txt")
data.tab
```

## IMPORT data from an EXCEL file XLS and XLSX formats

We will use the `readxl` package. Be sure to install this package `install.package("readxl")` and then load it using `library(readxl)`.

Here is code for reading in a XLS format

```{r}
data.xls <- read_excel("C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01.xls", sheet=1)
data.xls
```

Here is code for reading in a XLSX format

```{r}
data.xlsx <- read_excel("C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01.xlsx", sheet=1)
data.xlsx
```

## IMPORT data from a SPSS file in SAV format

In order to read datafiles from SPSS, other statistical software, and many other datafile types. To see detailed info look at help for the `foreign` package. Run the following code:

```
help(package = "foreign")
```
We will use this package to read in an SPSS datafile in the SAV format. The function we will use is `read.spss()`. See more detailed information on the various options by running `help(read.spss)`. For now we will specify the filename and tell the function to make sure the file is output as a data frame object.

```{r warning=FALSE, error=FALSE}
data.spss <- read.spss(file = "C:/MyGithub/A_Series_of_R_Workshops/datasets/Dataset_01.sav",
                       to.data.frame=TRUE)
data.spss
```

## Create some new variables and save the output

In the datafile we have weights measures at 2 time points and we have height. We can use this data to compute BMI. Since weight is in pounds and height is in inches, we can use the following formula:

`BMI_PRE=(WeightPRE*703)/((Height*12)*(Height*12))`

```{r}
data.csv$bmiPRE <- (data.csv$WeightPRE*703)/((data.csv$Height*12)**2)
```

And we'll do it again for the POST weights:

`BMI_POST=(WeightPOST*703)/((Height*12)*(Height*12))`

```{r}
data.csv$bmiPOST <- (data.csv$WeightPOST*703)/((data.csv$Height*12)**2)
```

## EXPORT or SAVE the updated data

We can save it out as a RData file using the `save()` function.

```{r}
save(data.csv, 
     file="C:/MyGithub/A_Series_of_R_Workshops/datasets/datacsv.RData")
```

Save the data out in a delimited format. First we'll do a comma delimited CSV file using `write.csv()`.

```{r}
write.csv(data.csv, 
          file="C:/MyGithub/A_Series_of_R_Workshops/datasets/datacsv.csv")
```

Next we'll do a TAB delimited text file using the `write.table()`.

```{r}
write.table(data.csv, 
          file="C:/MyGithub/A_Series_of_R_Workshops/datasets/datacsv.txt",
          sep="\t")
```




